---
title: "Exercise 3 --- Point Patterns in Space and Space-Time"
subtitle: "BSTA 374 · 25 points · Due Sunday, October 4, 2026, 11:59 pm"
author: "Eric M. Delmelle · Lehigh University"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
  pdf:
    papersize: letter
    geometry:
      - margin=1in
    toc: false
    colorlinks: true
    code-overflow: wrap
prefer-html: true
execute:
  eval: true
  echo: true
---

## What this exercise is

The largest lab of the semester. You will decide whether a set of case locations is clustered, at what spatial
scale, where the intensity is highest, and whether the clustering is also
concentrated in *time* --- which is a different claim, and the one that matters
for an outbreak.

**Data.** Two files, posted with this exercise. Put them in a `data/` folder
beside your `.qmd`.

| File | What it holds |
|---|---|
| `dengue_cases_sim.csv` | 500 simulated dengue cases: `case_id`, `onset_date`, `report_date`, `neighborhood`, `age_group`, `lon`, `lat` |
| `study_area.geojson` | the study-area boundary polygon |

These are **simulated** data modelled on a dengue surveillance setting. Nobody's
health information is involved, and the generating process is known --- which
means your conclusions can be checked.

Each part ends with **starting code**: copy it, fill in the blanks marked
`______`, and adapt. It is scaffolding, not a solution.

::: {.callout-important}
## Submission
Two things. Submit `LASTNAME_FI_Exercise3.qmd` --- the source --- to Course Site,
with your name in the document header. And publish the rendered HTML to your
semester GitHub repository, then paste the live URL into the Course Site
submission --- Part H is that second step, and it is the one that carries the
point.

**Due Wednesday, October 1, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Restart R and render from a clean session before submitting. Several parts
involve simulation --- set a seed at the top and use it, or your numbers will
change between the render you checked and the render I grade.
:::

```{r}
#| label: setup


library(sf)
library(dplyr)
library(ggplot2)
library(spatstat)
library(splancs)      # space-time K function
library(plotly)       # space-time cube

set.seed(800074585)      # your student ID works
```

---

## Part A --- From CSV to `ppp` (3 points)

Everything after this part runs on a `ppp` object, so it is worth knowing what
one is. `spatstat` does not work with `sf`. A **`ppp`** ("planar point pattern")
holds two things: the coordinates of the events, and a **window** --- the region
within which they could have been observed. Both are required, because every
statistic in this lab is a comparison against what would be expected inside that
window: the intensity is $n$ divided by its area, the expected nearest-neighbour
distance depends on that intensity, and the edge corrections depend on its shape.
Give `spatstat` the wrong window and every number below changes without any
warning.

1. Read the case file and the boundary. Report how many cases you have and what
   one row represents.
2. Convert the cases to an `sf` object in **EPSG:4326**, then project both the
   cases and the boundary to **EPSG:32618** (UTM 18N, metres). One sentence on
   why the analysis cannot be done in degrees.
3. Build a `ppp` whose window is the study-area polygon, then print it. Report
   the number of events and the intensity in cases per km².
4. Replace the polygon window with the bounding box of the cases
   (`Window(pp) <- owin(range(x), range(y))`) and recompute the intensity. Two or
   three sentences on how much it changed and why the choice of window is a
   modelling decision rather than a technicality. Then go back to the polygon.
5. Map the cases over the boundary.

### Starting code

```{r}
#| label: a-starter

cases_raw <- read.csv("data/dengue_cases_sim.csv", stringsAsFactors = FALSE)
area      <- st_read("data/study_area.geojson", quiet = TRUE) |> st_transform(32618)

cases <- cases_raw |>
  mutate(onset_date  = as.Date(onset_date),
         report_date = as.Date(report_date)) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(32618)

win <- as.owin(st_geometry(area))       # the observation window
pp  <- as.ppp(st_geometry(cases))       # coordinates only, with a bounding box

x <- pp$x # Save coordinates
y <- pp$y

Window(pp) <- owin(range(x), range(y)) # Replace polygon with rectangular bounding box

pp
intensity(pp) * 1e6

Window(pp) <- win                       # attach the real window

pp                                      # print it: n, window type, extent
npoints(pp)
intensity(pp) * 1e6                     # cases per km^2


#| label: part-a-map

ggplot() +
  geom_sf(
    data = area,
    fill = "white",
    color = "black",
    linewidth = 0.6
  ) +
  geom_sf(
    data = cases,
    size = 1,
    alpha = 0.6
  ) +
  labs(
    title = "Simulated Dengue Cases",
    x = "Easting (m)",
    y = "Northing (m)"
  ) +
  theme_minimal()
```
1.
The dataset contains 500 dengue cases, with each row representing one simulated dengue case and its associated information, including location and dates.

2.
The analysis should not be performed in degrees because degrees are angular units and do not provide consistent measurements of distance or area, while UTM coordinates are measured in meters.

3.
npoints(pp)
intensity(pp) * 1e6 

The point pattern contains 500 dengue case events. Using the study-area polygon as the observation window, the estimated intensity is approximately 2.63 cases per km.

4.
Window(pp) <- owin(range(x), range(y)) # Replace polygon with rectangular bounding box

pp
intensity(pp) * 1e6

Replacing the study-area polygon with the cases' bounding box decreases the intensity from approximately 2.63 to 2.35 cases per km², a decrease of about 10.6%. The rectangular bounding box includes space outside the actual study area, increasing the assumed observation area and lowering the calculated intensity. The window is a modelling decision because it represents where cases could have been observed; changing that region changes the assumptions and resulting spatial statistics.

---

## Part B --- Quadrat analysis (3 points)

1. Lay a 6 × 6 grid over the window, map the counts, and compute the mean,
   variance and **variance-to-mean ratio** from the counts by hand --- not from a
   canned function. Show the arithmetic.
2. State what the VMR is being compared against, and why: under complete spatial
   randomness the counts are Poisson, and a Poisson distribution has variance
   equal to its mean. Report where your value sits relative to 1.
3. Repeat the VMR for at least four grid resolutions and plot VMR against cells
   per side. Three or four sentences: what does the shape of that curve say about
   the claim "this pattern is clustered"?

### Starting code

```{r}
#| label: b-starter

qc  <- quadratcount(pp, nx = 6, ny = 6)
cnt <- as.vector(qc)

k <- length(cnt)

mean_cnt <- sum(cnt) / k
var_cnt  <- sum((cnt - mean_cnt)^2) / (k - 1)
VMR      <- var_cnt / mean_cnt

c(k = k, mean = mean_cnt, var = var_cnt, VMR = VMR)

grid_n <- c(3, 4, 6, 8, 12)
vmr <- sapply(grid_n, function(k) {
  v <- as.vector(quadratcount(pp, nx = k, ny = k))
  
  m <- sum(v) / length(v)
  s2 <- sum((v - m)^2) / (length(v) - 1)
  
  s2 / m

})
vmr

# Plot VMR against grid resolution
plot(
  grid_n, vmr,
  type = "b",
  pch = 19,
  xlab = "Cells per side",
  ylab = "Variance-to-Mean Ratio",
  main = "VMR Across Quadrat Resolutions"
)

abline(h = 1, lty = 2)
```
2.

c(k = k, mean = mean_cnt, var = var_cnt, VMR = VMR)
       k      mean       var       VMR 
 36.00000  13.88889 581.70159  41.88251 


For the 6x6 grid, there were 36 quadrats containing a total of 500 cases. The mean number of cases per quadrat was 13.89, while the variance was approximately 581.70. Dividing the variance by the mean gives a variance-to-mean ratio of approximately 41.88.

Under complete spatial randomness, quadrat counts are expected to follow a Poisson distribution. For a Poisson distribution, the variance is equal to the mean, so the expected VMR is 1. Our VMR of 41.88 is much greater than 1, indicating more variation among quadrats than would be expected under complete spatial randomness and providing evidence of spatial clustering.

3.

c(k = k, mean = mean_cnt, var = var_cnt, VMR = VMR)

grid_n <- c(3, 4, 6, 8, 12)
vmr <- sapply(grid_n, function(k) {
  v <- as.vector(quadratcount(pp, nx = k, ny = k))
  
  m <- sum(v) / length(v)
  s2 <- sum((v - m)^2) / (length(v) - 1)
  
  s2 / m

})
vmr
[1] 32.99900 35.17227 41.88251 32.31300 29.96188

The VMR remains much greater than 1 at every grid resolution examined, ranging from approximately 30.23 to 41.88. It increases from the 3x3 grid to a maximum at the 6x6 grid and then decreases for the finer 8x8 and 12x12 grids. Because the VMR remains well above 1 across all of the grid sizes, the conclusion that the cases are clustered is not dependent on a single grid resolution. However, the changing magnitude of the VMR shows that the measured strength of clustering depends on the spatial scale used for the analysis.
---

## Part C --- Nearest-neighbour statistics (3 points)

1. Compute `nndist()`, report the observed mean, and compute the CSR expectation
   $E(d) = 1/(2\sqrt{\lambda})$ **by hand** from your intensity. Report the
   Clark--Evans ratio $R$ and state the direction of the departure.
2. Run `clarkevans.test()` with a Monte Carlo null. The window is an irregular
   polygon, so the Donnelly correction is unavailable --- use
   `correction = "cdf"` with `method = "MonteCarlo"`. One sentence on why an edge
   correction is needed at all.
3. Plot the G function with a simulation envelope. Two or three sentences on what
   G shows that the single ratio $R$ does not.

### Starting code

```{r}
#| label: c-starter

nnd    <- nndist(pp)
lambda <- intensity(pp)
d_exp <- 1 / (2 * sqrt(lambda))                      # 1 / (2 sqrt(lambda))

R <- mean(nnd) / d_exp

c(observed = mean(nnd), expected = d_exp, R = R)

clarkevans.test(pp, correction = "cdf", alternative = "clustered",
                method = "MonteCarlo", nsim = 99)

plot(envelope(pp, Gest, nsim = 39, verbose = FALSE), main = "", legend = FALSE)
```
1.
c(observed = mean(nnd), expected = d_exp, R = R)
   observed    expected           R 
215.5165165 308.5023198   0.6985896 

The observed mean nearest-neighbor distance was approximately 215.52 m. Using the study-area intensity, the expected nearest-neighbor distance under complete spatial randomness was approximately 308.50 m. Therefore, the Clark–Evans ratio was 0.699.

2.
clarkevans.test(pp, correction = "cdf", alternative = "clustered",
                method = "MonteCarlo", nsim = 99)
              
data:  pp
R = 0.65464, p-value = 0.01
alternative hypothesis: clustered (R < 1)

Because (R<1), the observed cases are closer together than expected under complete spatial randomness, indicating a departure in the direction of spatial clustering.

An edge correction is needed because points near the study-area boundary may have nearest neighbors outside the observation window that cannot be observed, which can bias nearest-neighbor distances upward.

3.
The G function examines the full distribution of nearest-neighbor distances rather than summarizing the pattern with one value like the Clark–Evans ratio. It shows the spatial distances at which the observed pattern differs from complete spatial randomness; for example, an observed G curve above the simulation envelope at short distances indicates more close neighbors than expected under CSR. Therefore, G can reveal the scale of clustering, while R only gives an overall measure of whether nearest neighbors tend to be closer or farther apart than expected.
---

## Part D --- Ripley's K across scales (4 points)

1. Estimate $K(r)$ with an edge correction, then plot the centred L function
   $\hat{L}(r) - r$ against $r$.
2. Add a **simultaneous** envelope (`global = TRUE`) from at least 99 CSR
   simulations. State the distances over which the curve leaves the band.
3. Read a **characteristic scale** off the curve --- the $r$ at which the
   departure peaks --- and translate it into something a public health officer
   would recognise: a few city blocks, a neighbourhood, half the city.
4. Three or four sentences: $K$ assumes the intensity is constant across the
   window. Looking at your map from Part A, is that plausible here? Say what the
   consequence is for reading the curve --- what else, besides interaction
   between cases, could put it above the envelope.

::: {.callout-warning}
`global = TRUE` with `nsim = 99` takes a minute or two. Develop with
`nsim = 19`, raise it once, and do not re-render in a loop.
:::

### Starting code

```{r}
#| label: d-starter

L <- Lest(pp, correction = "isotropic")
plot(L, . - r ~ r, main = "Centred L Function", xlab = "Distance r (m)", ylab = "L(r) - r", legend = FALSE)

env <- envelope(pp, Lest, nsim = 99, correction = "isotropic",
                global = TRUE, verbose = FALSE)
plot(env, . - r ~ r, main = "Centred L Function with Global CSR Envelope", xlab = "Distance r (m)", ylab = "L(r) - r", legend = FALSE)

abline(h = 0, lty = 2)

outside <- env$obs > env$hi | env$obs < env$lo

range(env$r[outside])

departure <- env$obs - env$r

peak_index <- which.max(departure)
peak_r <- env$r[peak_index]
peak_departure <- departure[peak_index]

c(
  peak_distance_m = peak_r,
  peak_departure = peak_departure
)
```
2.
range(env$r[outside])
[1]   14.64052 3747.97331

The observed centered L curve rises above the simultaneous CSR envelope at approximately 14.6 m and remains above the envelope through approximately 3748 m. This indicates evidence of clustering over a broad range of spatial distances rather than at only one particular distance.

3.
c(
  peak_distance_m = peak_r,
  peak_departure = peak_departure
  
peak_distance_m  peak_departure 
       790.5881       1566.7923
       
The largest positive departure occurs at approximately 790.6 m, or about 0.79 km. From a public health perspective, this represents roughly several city blocks or the scale of a small neighborhood. This suggests that the strongest spatial concentration of dengue cases occurs at approximately the neighborhood scale

4.
The ordinary K-function assumes that the underlying intensity is constant throughout the study area. Based on the Part A map, this assumption does not appear very plausible because some areas have visibly higher concentrations of cases than others. Therefore, the curve being above the CSR envelope could result not only from interaction or dependence between nearby cases, but also from spatial variation in underlying risk, such as differences in population density, mosquito habitat, environmental conditions, or neighbourhood characteristics. Thus, the K-function provides evidence of spatial aggregation but does not by itself show that nearby cases directly influence one another.
---

## Part E --- Kernel density estimation and bandwidth choice (4 points)

K tells you *whether* and *at what distance*; it does not tell you *where*. That
is what a kernel intensity surface is for.

1. Estimate the intensity with `density.ppp()` and convert to **cases per km²**.
   Map it with the cases drawn on top.
2. Produce the same surface at **three bandwidths** --- one clearly too small,
   one clearly too large, and one you would defend --- on a **common colour
   scale** so the panels are comparable. Two or three sentences on what changes
   and what does not.
3. Compare at least two automatic selectors (`bw.diggle()`, `bw.ppl()`,
   `bw.scott()`). Report the values in metres, say which you adopt and why, and
   note whether your answer to step 2 would change if you adopted a different one.
4. Three or four sentences on what this map is and is not. Where is the intensity
   highest? What would you need in order to turn this surface into a statement
   about **risk** rather than about counts, and what is the name of the quantity
   you would compute instead?

::: {.callout-tip}
## What earns full marks here
Step 4, and the common colour scale in step 2. Two surfaces at different
bandwidths with independently stretched legends look like different data; a
reader cannot compare them, and neither can you.
:::

### Starting code

```{r}
#| label: e-starter

sig <- bw.diggle(pp)
c(diggle = bw.diggle(pp), ppl = bw.ppl(pp), scott = mean(bw.scott(pp)))

kde <- density(pp, sigma = sig, eps = 50) * 1e6       # cases per km^2
plot(kde, main = "", ribargs = list(las = 1))
plot(Window(pp), add = TRUE, border = "grey40")
points(pp, pch = 16, cex = 0.25, col = rgb(0, 0, 0, 0.35))

sigmas <- c(sig / 3, sig, sig * 3)                   # too small, defensible, too large
surf   <- lapply(sigmas, function(s) density(pp, sigma = s, eps = 50) * 1e6)
zmax   <- max(sapply(surf, max))                      # one scale for all panels


par(mfrow = c(1, 3))

for (i in seq_along(surf)) {
  plot(
    surf[[i]],
    main = paste0("sigma = ", round(sigmas[i]), " m"),
    zlim = c(0, zmax),
    ribargs = list(las = 1)
  )
  
  plot(Window(pp), add = TRUE, border = "grey40")
  points(pp, pch = 16, cex = 0.2, col = rgb(0, 0, 0, 0.35))
}

par(mfrow = c(1, 1))
```
2.
The smallest bandwidth produces a very detailed and irregular surface with many narrow local peaks around individual cases, suggesting that it is undersmoothed. The largest bandwidth produces a much smoother surface in which nearby concentrations merge together and important local variation is lost. The intermediate bandwidth preserves the main areas of high intensity while smoothing out much of the point-level noise, making it the most useful representation of the overall spatial pattern. Although the amount of detail changes across bandwidths, the major areas of high and low case intensity remain generally similar.

3.
c(diggle = bw.diggle(pp), ppl = bw.ppl(pp), scott = mean(bw.scott(pp)))
   diggle       ppl     scott 
 168.6955  469.9289 1250.5481 
 
The automatic bandwidth selectors produced estimates of 168.70 m for bw.diggle(), 469.93 m for bw.ppl(), and 1250.55 m for bw.scott(). I adopted the PPL bandwidth of 469.93 m because it provides a middle level of smoothing: the Diggle bandwidth produces a much more locally variable surface, while the Scott bandwidth smooths over a much larger area and may hide important local concentrations. The PPL bandwidth therefore provides a reasonable balance between preserving meaningful spatial structure and avoiding excessive small-scale variation. Using one of the other selectors would change the amount of smoothing and the exact peak intensity, but it would not necessarily change the main conclusion about the general locations of high and low case intensity. My answer to Step 2 would not  change if I adopted a different automatic selector: the Diggle bandwidth would emphasize more small-scale peaks and the Scott bandwidth would produce broader, smoother regions, but the overall locations of relatively high and low case intensity would remain similar.

4.
The kernel density map shows the estimated intensity of dengue cases, measured in cases per km², rather than the risk of infection. The highest intensity occurs in the Aguablanca area, where the cases are most densely concentrated. To make a statement about dengue risk, we would also need information on the underlying population at risk, such as population density or the number of residents in each area. We could then compare case intensity with population intensity to estimate a relative risk surface rather than simply mapping case counts.
---

## Part F --- Space-time kernel density (3 points)

A single intensity surface pools six months. An outbreak is a hotspot that
exists in some weeks and not others, and pooling hides exactly that.

1. Split the cases into **monthly** patterns and estimate a kernel surface for
   each, using **one bandwidth for all months** and **one colour scale for all
   months**. Display them as small multiples.
2. Two or three sentences on why both of those constraints are necessary, and
   what a reader would wrongly conclude if each panel were scaled to its own
   maximum.
3. Name the months and locations where the intensity is high, and say which
   hotspots persist across months and which appear and disappear.
4. Build a **space-time cube**: an interactive 3D scatter with easting, northing
   and day-of-study on the three axes, coloured by neighbourhood. Rotate it until
   the structure is visible, then set a viewing angle that makes your point and
   keep it. One or two sentences on what the cube shows that the small multiples
   do not, and vice versa --- look in particular for the difference between a
   compact blob and a vertical column.

::: {.callout-warning}
## plotly renders in HTML only
The cube is an HTML widget: it appears in an HTML render and is silently dropped
from a PDF one. The two chunks below are gated on `knitr::is_html_output()`, so
whichever format you render gets a working figure. If you publish the HTML, keep
`embed-resources: true` --- otherwise the widget lives in a `_files/` folder that
has to travel with the page.
:::

::: {.callout-note}
## If you want the proper estimator
Steps 1--3 are a discretised space-time KDE: one spatial bandwidth, and a
temporal "bandwidth" equal to one month, applied as a hard cut. The continuous
version smooths in time as well, with a separate temporal bandwidth, and lives
in the `sparr` package --- `spattemp.density()` with `h` and `lambda`, then
`spattemp.slice()` to pull out time slices. Using it is worth up to **2 bonus
points** if the monthly version is already complete and correct. Do not start
here.
:::

### Starting code

```{r}
#| label: f-starter

cases$month <- format(cases$onset_date, "%Y-%m")

month_levels <- sort(unique(cases$month))
h <- 469.9289                                  # one bandwidth, chosen in Part E

by_month <- lapply(month_levels, function(m) {
  sub <- cases[cases$month == m, ]

  xy_sub <- st_coordinates(sub)

  p <- ppp(
    xy_sub[, 1],
    xy_sub[, 2],
    window = Window(pp)
  )
  density(p, sigma = h, eps = 50) * 1e6
})

zmax <- max(sapply(by_month, max))        # common scale across months

par(mfrow = c(2, 3), mar = c(1, 1, 2, 2))

for (i in seq_along(by_month)) {

  plot(
    by_month[[i]], 
    main = month_levels[i],
    zlim = c(0, zmax),     
    ribargs = list(las = 1))

  sub <- cases[cases$month == month_levels[i], ]
  xy_sub <- st_coordinates(sub)

  points(
    xy_sub[, 1],
    xy_sub[, 2],
    pch = 16,
    cex = 0.25
  )
}

par(mfrow = c(1, 1))


xy <- st_coordinates(cases)

cube <- data.frame(
  E = xy[, 1],
  N = xy[, 2],
  day = as.numeric(
    cases$onset_date - min(cases$onset_date)
  ),
  hood = cases$neighborhood
)

# BONUS
library(sparr)

tt <- as.numeric(cases$onset_date - min(cases$onset_date))

h_bonus <- 469.9289

st_bw <- OS.spattemp(pp, tt = tt)

st_bw

lambda_bonus <- st_bw["lambda"]

lambda_bonus

st_kde <- spattemp.density(
  pp = pp,
  tt = tt,
  h = h_bonus,
  lambda = lambda_bonus,
  tlim = range(tt),
  sres = 128,
  verbose = FALSE
)

st_kde

slice_dates <- as.Date(c(
  "2024-01-15",
  "2024-02-15",
  "2024-03-15",
  "2024-04-15",
  "2024-05-15",
  "2024-06-15"
))

slice_days <- as.numeric(
  slice_dates - min(cases$onset_date)
)

st_slices <- spattemp.slice(
  st_kde,
  tt = slice_days
)

st_slices

zmax_bonus <- max(
  sapply(st_slices$z.cond, function(z) max(z$v, na.rm = TRUE))
)

par(mfrow = c(2, 3), mar = c(1, 1, 2, 2))

for (i in seq_along(st_slices$z.cond)) {
  plot(
    st_slices$z.cond[[i]],
    main = format(slice_dates[i], "%Y-%m"),
    zlim = c(0, zmax_bonus),
    ribargs = list(las = 1)
  )
}

par(mfrow = c(1, 1))
```

```{r}
#| label: f-cube-html
#| eval: !expr knitr::is_html_output()

plotly::plot_ly(
  cube,
  x = ~E,
  y = ~N,
  z = ~day,
  color = ~hood,
  type = "scatter3d",
  mode = "markers",
  marker = list(
    size = 3,
    opacity = 0.75
  )
) |>
  plotly::layout(
    scene = list(
      xaxis = list(title = "Easting (m)"),
      yaxis = list(title = "Northing (m)"),
      zaxis = list(title = "Day of study"),
      camera = list(
        eye = list(x = 1.6, y = 1.6, z = 0.9)
      )
    )
  )
```

```{r}
#| label: f-cube-static
#| eval: !expr isFALSE(knitr::is_html_output())

scatterplot3d::scatterplot3d(cube$E, cube$N, cube$day, pch = 16, cex.symbols = 0.4,
                             color = as.integer(factor(cube$hood)),
                             xlab = "easting (m)", ylab = "northing (m)", zlab = "day")
```

2.
Using the same bandwidth ensures that differences among months reflect real changes in the spatial distribution of cases rather than different amounts of smoothing. Using one colour scale ensures that the same colour represents the same case intensity in every month. If each panel were scaled to its own maximum, months with relatively weak concentrations could appear just as intense as months with much stronger hotspots.

3.
The clearest high-intensity hotspots occur in Floralia in February, Aguablanca in March, and Siloe in May. January, April, and June have much lower peak intensities on the common scale and do not show hotspots as strong as those three months. The strong Floralia, Aguablanca, and Siloe hotspots therefore appear for limited periods and then disappear, while El Retiro shows cases across much of the study period, representing a more persistent but generally lower-intensity concentration.

4.
The space-time cube shows whether cases at the same location occur during a short period or persist over many days. The strong monthly hotspots appear more like compact blobs, representing many cases concentrated in both space and time, whereas El Retiro forms more of a vertical column, showing cases occurring in roughly the same location throughout much of the study period. The monthly KDE maps make it easier to compare the geographic intensity of hotspots month by month, while the cube makes their duration and temporal continuity easier to see.

Bonus.
The continuous space-time KDE used a spatial bandwidth of 469.93 m and a temporal bandwidth of 15.26 days. Unlike the monthly analysis, which places cases into discrete one-month intervals, this estimator smooths continuously through time, so cases near the end of one month also influence nearby dates in the next month. The resulting slices show the emergence and disappearance of spatial concentrations more gradually and avoid artificial breaks at calendar-month boundaries. This provides a more continuous representation of how dengue case intensity changes jointly across space and time.
---

## Part G --- The space-time K function (4 points)

Spatial clustering and space-time clustering are different claims. A
neighbourhood with chronically high incidence is clustered in space in every
month; an outbreak is clustered in space *and* time together. Diggle's space-time
K function separates them by comparing the joint function with the product of the
two marginals:

$$
D(s, t) = K_{st}(s, t) - K_s(s)\,K_t(t)
$$

$D = 0$ means space and time are independent --- cases may cluster spatially,
temporally, or both, but not *jointly*. $D > 0$ is space-time interaction.

1. Build the inputs `splancs` expects: a two-column matrix of projected
   coordinates, a numeric time vector (days since the first onset), and the
   boundary as a matrix of vertices.
2. Compute $K_s$, $K_t$ and $K_{st}$ with `stkhat()`. Justify your choice of
   spatial distances `s` and time lags `tm` in one sentence each --- what scales
   are plausible for a mosquito-borne disease?
3. Plot **both** $D(s,t)$ and the standardised version
   $D_0(s,t) = K_{st}/(K_s K_t) - 1$. One or two sentences on why the two
   surfaces have different shapes: $K$ functions are cumulative, so $D$
   accumulates and then flattens, while $D_0$ measures the *proportional* excess
   and falls away once you leave the scale of the clustering. This is the same
   move as plotting $\hat{L}(r) - r$ instead of $\hat{K}(r)$ in Part D.
4. A surface is hard to read against nothing. Build a **permutation envelope**:
   shuffle the onset dates across the fixed locations, recompute $D_0$, repeat 99
   times, and take the upper 97.5% quantile at each $(s, t)$. Then plot $D_0$
   against $s$ as a curve per time lag, with its envelope --- the space-time
   analogue of the envelope you used for $L$. Shuffling the dates destroys any
   space-time interaction while leaving both marginal patterns exactly as
   observed, which is the correct null here.
5. Read the peak off those curves --- a distance and a lag --- and say what the
   pair means for this disease.
6. Run `stmctest()` for a single global p-value. It sums the residuals
   $\hat{D}(s,t)$ over the whole grid into one statistic $U$, recomputes it under
   permuted dates, and returns the observed value as `$t0` and the permuted ones
   as `$t`. Report the p-value as
   $(1 + \#\{t \geq t_0\}) / (1 + B)$ and plot the null distribution with the
   observed value marked. One sentence on what this test adds to the curves in
   step 4, and what it cannot tell you.
7. Five or six sentences. Tabulate cases by neighbourhood and month. At least one
   neighbourhood contributes to the spatial clustering of Part D without
   contributing to the space-time interaction here. Identify it, explain what
   distinguishes it from the others, and say what each of the two patterns would
   mean for someone deciding where to send a vector-control team this week.

::: {.callout-warning}
## Never report $p = 0$
`mean(mc$t >= mc$t0)` returns 0 when no permutation beats the observed value,
and 0 is not a p-value --- with $B$ permutations you cannot resolve anything
below $1/(B+1)$. Always add the observed value to both counts, as above. With
999 permutations the smallest reportable value is 0.001, and the honest phrasing
is "$p \leq 0.001$", not "$p < 0.0001$".
:::

::: {.callout-note}
`splancs` is old and its arguments are terse: `stkhat(pts, times, poly, tlimits,
s, tm)`, where `tlimits` is the time range, `s` the spatial distances and `tm`
the time lags. `?stkhat` and `?stmctest` are short. A **Knox test** on a distance
and time threshold of your choosing is an acceptable substitute if you defend
both thresholds --- say explicitly if you go that route.
:::

### Starting code

```{r}
#| label: g-starter

xy    <- st_coordinates(cases)
pts   <- as.points(xy[, 1], xy[, 2])
times <- as.numeric(cases$onset_date - min(cases$onset_date))
bdry  <- st_coordinates(area)[, 1:2]         # drop the closing vertex if duplicated

if (all(bdry[1, ] == bdry[nrow(bdry), ])) {
  bdry <- bdry[-nrow(bdry), ]
}

s  <- c(
  seq(50, 1000, by = 50), 
  seq(1200, 3000, by = 200)
  )   # finer near the origin

tm <- seq(3, 70, by = 3)                     # days

stk <- stkhat(
  pts, 
  times, 
  bdry, 
  c(0, max(times)), 
  s, 
  tm
  )

D  <- stk$kst - outer(stk$ks, stk$kt)                  # raw excess
D0 <- stk$kst / outer(stk$ks, stk$kt) - 1              # standardised excess

par(mfrow = c(1, 2))

persp(
  s, tm, D,  
  theta = 30, 
  phi = 25,
  xlab = "distance (m)", 
  ylab = "lag (days)", 
  zlab = "D"
)

persp(
  s, tm, D0, 
  theta = 30, 
  phi = 25,
  xlab = "distance (m)", 
  ylab = "lag (days)", 
  zlab = "D0"
)

par(mfrow = c(1, 1))

# permutation envelope: shuffle the dates, keep the locations
B <- 99
perm <- replicate(B, {
  st <- stkhat(pts, sample(times), bdry, c(0, max(times)), s, tm)
  
  st$kst / outer(st$ks, st$kt) - 1
})

hi <- apply(perm, c(1, 2), quantile, 0.975)

k <- c(1, 3, 7, 14) # lags to show

matplot(s, D0[, k], type = "l", lty = 1, lwd = 2, xlab = "distance s (m)",
        ylab = expression(D[0](s, t)))

matlines(s, hi[, k], lty = 3, lwd = 1)

abline(h = 0, col = "grey")

legend("topright", legend = paste("lag <=", tm[k], "d"), lty = 1, lwd = 2,
       col = 1:4, bty = "n")

D0_peak <- D0
D0_peak[!is.finite(D0_peak)] <- NA

ij <- which(
  D0_peak == max(D0_peak, na.rm = TRUE),
  arr.ind = TRUE
)[1, ]

c(
  peak_distance_m = s[ij[1]],
  peak_lag_days = tm[ij[2]],
  peak_D0 = D0_peak[ij[1], ij[2]]
)

mc <- stmctest(pts, times, bdry, c(0, max(times)), s, tm, nsim = 999, quiet = TRUE)
p_global <- (
(1 + sum(mc$t >= mc$t0)) / (1 + length(mc$t))
) # global Monte Carlo p-value

p_global

hist(mc$t, xlim = range(c(mc$t, mc$t0)), xlab = "U under permuted dates", main = "")
abline(v = mc$t0, col = "firebrick", lwd = 2)

table(cases$neighborhood, cases$month)
```
2.
Spatial distances from 50 m to 3000 m were used because dengue transmission can operate at very local household or block scales while broader neighborhood-scale patterns may extend over several kilometres. Time lags from 3 to 70 days were used to capture short-term clustering over days to several weeks, while still allowing the analysis to detect whether cases remain associated over longer periods.

3.
Both D(s,t) and the standardized D_0(s,t) measure departures from independence between space and time, with positive values indicating excess space-time interaction. The raw D surface tends to accumulate as spatial distance and time lag increase because K-functions are cumulative, so it often rises and then flattens. In contrast, D_0 represents the proportional excess over what would be expected from the spatial and temporal marginal patterns, making the characteristic scale of the interaction easier to identify.

4.
The permutation envelope was generated by shuffling onset dates across the fixed case locations 99 times. This preserves the observed spatial clustering and the observed temporal distribution while destroying any association between a particular location and a particular onset time. Therefore, values of the observed D_0 curve above the 97.5% permutation envelope indicate space-time interaction beyond what would be expected from the spatial and temporal patterns separately.

5.
peak_distance_m   peak_lag_days         peak_D0 
      50.000000        3.000000        5.702277 
The strongest space-time interaction occurred at a spatial distance of 50 m and a time lag of 3 days, with D_0 ~ 5.70. This means cases occurring very close together geographically and within a few days of one another occurred much more often than expected if space and time were independent. For dengue, this pattern is consistent with very localized, short-term clustering that could occur around nearby households or a small group of city blocks. Because 50 m and 3 days are the smallest values tested, the true characteristic scale may actually be smaller than the resolution of this analysis.

6.
p_global
[1] 0.001
The global Monte Carlo test provided strong evidence of overall space-time interaction, with p < 0.001. This means the joint clustering of cases in space and time is stronger than would be expected if onset dates were randomly assigned to the observed case locations. The global test provides a single overall significance test for space-time interaction, whereas the D_0 curves show the particular spatial distances and time lags at which that interaction occurs. However, the global test cannot identify the scale at which the clustering is strongest.

7.
             
              2024-01 2024-02 2024-03 2024-04 2024-05 2024-06
  Aguablanca        5       7     105       5       1       2
  El Retiro        20      13      24      14      13      13
  Floralia          3      72       6       4       2       8
  La Base           3       8       1       4       4       1
  Melendez          7       4       3       3       4       2
  San Nicolas       4       4       1       5       2       5
  Siloe             6       3       3       2      84       3
  Terron            2       3       2       7       0       3
  
The neighborhood-by-month table shows that El Retiro is the clearest example of an area contributing to spatial clustering without producing the same type of strong space-time interaction as the outbreak hotspots. El Retiro has cases consistently throughout the six-month study, with monthly counts ranging from 13 to 24, so its cases form a persistent spatial concentration rather than a short-lived temporal outbreak. In contrast, Floralia has 72 cases in February, Aguablanca has 105 cases in March, and Siloe has 84 cases in May, creating strong concentrations in both space and time. For a vector-control team, persistent clustering in El Retiro could indicate an area with chronically elevated dengue activity that may justify ongoing prevention and surveillance. The concentrated peaks in Floralia, Aguablanca, and Siloe would instead indicate locations where a rapid, targeted response may be especially important during the specific weeks when cases surge. Thus, spatial K identifies areas with consistently high clustering, while the space-time K analysis helps distinguish those persistent areas from active outbreak clusters.
---

## Part H --- Rendering and publication (1 point)

The `.qmd` is the submission, but the rendered page is what anyone else can
actually read --- and by the end of the semester this repository is your
portfolio.

1. Render to **HTML** from a clean session, with the seed set and every package
   loaded explicitly, then commit both the `.qmd` and the rendered output to your
   semester repository, in the folder your site is built from.
2. Push, wait for GitHub Pages to rebuild, and **open the published URL yourself**
   to check it. A page that renders locally and 404s online is not submitted.
3. Put the URL of the published page in your document header and in your Course
   Site submission.

The point is for a page that works: rendered, pushed, and reachable at the URL
you supply.

::: {.callout-warning}
## Two things that break the published page
`embed-resources: true` is already set in the header --- keep it. Without it the
`plotly` cube in Part F lives in a `_files/` folder that must be committed
alongside the HTML, and forgetting it gives a page with a blank space where the
cube should be.

Relative paths to `data/` work locally because your working directory is the
`.qmd`'s folder. They will also work on the site *as long as you commit the data
files*. Check the live page, not the local preview.
:::

---

## AI use {.unnumbered}

End your document with a short **AI use note**: which tools, what for, at which
stage. Two or three sentences.

AI use is unrestricted in this course and costs you nothing, but you must be able
to explain every line you hand in. The note carries no points --- it is simply
expected, the way a methods section is.

AI was used in the improving of interpretations and the debugging of code. ChatGPT was the only LLM used in this task. The AI was only used when needed.
---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct import, CRS and window; `ppp` built and explained; effect of the window argued | 3 |
| **B** | VMR computed from counts; benchmark of 1 explained; resolution sweep interpreted | 3 |
| **C** | $E(d)$ derived and $R$ interpreted; Monte Carlo test correct; G compared with $R$ | 3 |
| **D** | Centred L with a simultaneous envelope; characteristic scale read and translated; stationarity assumption discussed | 4 |
| **E** | Surface in interpretable units; three bandwidths on a common scale; selectors compared; density-is-not-risk argument | 4 |
| **F** | Monthly surfaces with common bandwidth and scale; persistent vs transient hotspots named; space-time cube | 3 |
| **G** | Space-time K computed; $D_0$ with a permutation envelope; peak distance and lag read off; the spatial-only neighbourhood identified and the operational difference explained | 4 |
| **H** | Clean render to HTML, published to GitHub Pages; working URL supplied | 1 |
| | **Total** | **25** |

Writing counts for roughly 15% within each part --- more than in Exercise 1,
because more of this lab is interpretation. A submission that is all code and no
sentences cannot score above about 50%.

---

## Getting help

- Office hours, or email with the code and the exact error message.
- `splancs` in Part G is the one place where the documentation is thin. Start
  early enough that you can ask.
- Stuck for more than thirty minutes on something mechanical: ask rather than
  persevere.
